Data Analysis Methods for the Detection of the Epoch of Reionization

### 2017-11-15 PAPER.tex
-Made Aaron's edits to Section 3.1.1 and 3.1.2

For our analysis, we use $8$ hours of LST (RA $0.5$-$8.6$ hours) and $51$ total baselines (\citet{ali_et_al2015} uses an RA range of $0$-$8.6$ hours). All power spectrum results are produced for a center frequency of $151$ MHz using a width of $10$ MHz ($20$ channels), identical to the analysis in \citet{ali_et_al2015}. We note that the PAPER-64 dataset that we use in this case study differs from that in \citet{ali_et_al2015} mainly by the fringe-rate filter. \cc{Danny says the next part is wrong... } In \citet{ali_et_al2015}, the applied filter was degraded by widening it in fringe-rate space. This was chosen in order to increase the number of independent modes and reduce signal loss. With the development of a robust method for assessing signal loss, we feel comfortable using a narrow filter --- the optimal fringe-rate filter --- in order to maximize sensitivity. This filter is computed for a fiducial $30$ m baseline at $150$ MHz, the center frequency in our band. The filter in both the fringe-rate domain and time domain is shown in Figure \ref{fig:frp}.

@@

We are interested in quantifying how much variance in $\textbf{e}$ is lost after weighting $\textbf{r}$ and estimating the power spectrum according to OQE formalism. We investigate this by comparing two quantities we define as the input power spectrum and output power spectrum: $P_{in}$ and $P_{out}$, defined as (ignoring normalization factors):

@@

However, deeper investigations of these terms reveal that they contain non-negligible power. In fact, foreground modes and signal modes are anti-correlated on average. This negative cross-term power arises from inverting a covariance matrix with a strong diagonal and small off-diagonal terms, as we expect to be the form for $\textbf{C}_{r}$. Through the inversion process, the off-diagonal terms become negative and ... This effect is derived mathematically and explained more in depth in \citet{switzer_et_al2015}.

@@

Our signal loss transfer function for one $k$ value is shown in Figure \ref{fig:sigloss_transfercurve} and compares the weighted case ($\textbf{R} = \textbf{C}^{-1}$, left plot) with an unweighted case ($\textbf{R} = \textbf{I}$, right plot). The figures depict a kernel density estimation of our input to output power spectra. To sample the distribution, we run $40$ total injection levels, where the amplitude of \textbf{e} is increased in even increments in log-space and ranges from $P_{in} \sim -10^{13} \, mK^{2} (h^{-1} Mpc)^{3}$ (negative $P_{in}$ is injected when one copy of \textbf{e} in Equation \eqref{eq:Pin} is negative and the other is positive) to $P_{in} \sim10^{13} \, mK^{2} (h^{-1} Mpc)^{3}$. For each injection, we perform $20$ bootstraps over baselines, as explained in Section \ref{sec:Boot}, resulting in a total of $800$ data points on our $P_{in}$ vs. $P_{out}$ grid. We have verified that the transfer function is symmetric for negative and positive $P_{in}$'s, and so we fold all the values into positive ones to increase our signal-to-noise.

The solid black diagonal line in Figure \ref{fig:sigloss_transfercurve} shows unity-transfer, which we expect to occur for the unweighted case ($\textbf{R} = \textbf{I}$). The solid grey horizontal lines mark the peak of the data distribution $\hat{\textbf{p}}$ (the distribution is created from the $20$ bootstraps) for both the inverse covariance weighted and unweighted cases. One can eyeball that a data value of $10^{5} \, mK^{2} (h^{-1} Mpc)^{3}$, for example, would map approximately to a value of $\sim10^{8} \, mK^{2} (h^{-1} Mpc)^{3}$ using the transfer curve for the inverse covariance weighted case, implying a signal loss factor of $\sim1000$.

\begin{figure*}
	\centering
	\includegraphics[width=1\textwidth]{plots/sigloss_transfercurve.png}
	\caption{Signal loss transfer functions showing the relationship of $P_{in}$ and $P_{out}$, as defined by Equations \eqref{eq:Pin} and \eqref{eq:sigloss}. Kernel density estimations of the transfer functions are shown as colored heat-maps for the cases of inverse covariance weighted PAPER-64 data (left) and unweighted data (right). The solid black diagonal line marks a perfect unity mapping, and the solid grey horizontal line denotes the peak of $\hat{\textbf{p}}$, the data distribution. From these plots, it is clear that inverse covariance weighting results in $\sim3$ orders of magnitude of signal loss for power spectrum values above $\sim$$2 \times 10^{4}$ $mK^{2} \, (h^{-1} \, Mpc)^{3}$, whereas the unweighted case does not exhibit loss.}
	\label{fig:sigloss_transfercurve}
\end{figure*}

One peculiar aspect of Figure \ref{fig:sigloss_transfercurve} is the fact that at low $P_{in}$ values it appears that we can have signal gain ($P_{out} > P_{in}$). This is unphysical in nature but caused due to the cross-terms dominating in Equation \eqref{eq:pout_expand} once $\textbf{e}$ becomes small. As $\textbf{e}$ increases, we move into a regime where $P_{out} \sim P_{in}$, and then eventually into a regime where $P_{out} < P_{in}$ when $\textbf{e}$ is large enough to be destroyed if weighting the data using itself. Although we only show figures for one $k$ value, we note that the shape of the transfer curve is nearly identical for all $k$'s (though we treat each $k$ separately).

Finally, it is worth thinking about the spread, or scatter, in our transfer curves in more detail. One might imagine that there should be one true $P_{out}$ value for every $P_{in}$, implying one well-determined signal loss factor for every data value. In practice we have found that this is not the case, as evidenced by the colored `heat-map' in Figure \ref{fig:sigloss_transfercurve}. The cause of this scatter can be thought of as comprised of two components.

The first component arises from the fact that the values of the cross-terms in Equation \eqref{eq:pout_expand} change for different mock EoR signals, $\textbf{e}$. Since we draw different random $\textbf{e}$'s for every bootstrap, there is a range of $P_{out}$ values that results from the same $P_{in}$. This randomness causes the spread seen in the unweighted case (right plot of Figure \ref{fig:sigloss_transfercurve}), as well as most of the spread in the weighted case (left plot).

The second component of scatter is less intuitive and only applies (at a low level) to the weighted case, but represents intrinsic scatter in signal loss. The origin of this scatter stems from how much the random EoR signal $\textbf{e}$ happens to look like the data $\textbf{x}$. For example, if we are extremely unlucky, one realization of $\textbf{e}$ would have the same shapes, or eigenvectors, as $\textbf{x}$. If weighting using empirically-derived covariances, this entire EoR signal would be destroyed. On the other hand, the less that $\textbf{e}$ looks like $\textbf{x}$, the less signal loss that would result. The intrinsic scatter we can get is not a dominant factor in this case but it is important to correct for the fact that a particular $P_{out}$ value could arise from a range of $P_{in}$ values.

@@

Beginning with smoothed probability distributions of the transfer function for the two cases (weighted and unweighted), we fit a polynomial to each, representing mean signal loss functions. We then account for the second component of scatter as described previously by de-convolving out $P_{out, unweighted}$ from $P_{out, weighted}$ for each $P_{in}$ value. In other words, we only account for the extra scatter that the weighted case has over the unweighted case, which, as can be seen in Figure \ref{fig:sigloss_transfercurve}, is not large.

Equipped with our new signal loss curves (a one-to-one mapping function for the unweighted case, and a somewhat smeared out function for the weighted case), we multiply our data distribution $\hat{\textbf{p}}$ element-wise to all $P_{in}$ columns, treating positive and negative halves of the data distribution separately. We note that for PAPER-64, $\hat{\textbf{p}}$ is sometimes narrowly peaked at positive power spectrum values (signifying detections of power), and sometimes centered around zero, representing measurements that are consistent with zero.

Finally, we sum up all rows of $P_{out}$ where each row is a $P_{in}$ probability distribution allowed by our measurements. This results in a final $P_{in}$ distribution --- the distribution of our data as seen through the signal loss lens. We show these final distributions in Figure \ref{fig:sigloss_datadist_inversecovariance} for every $k$. We compute power spectrum points from the peak of the histograms, and power spectrum errors from $95\%$ confidence intervals.

\begin{figure*}
	\centering
	\includegraphics[trim={1cm 0cm 1cm 1cm},width=1\textwidth]{plots/sigloss_datadist_inversecovariance.png}
	\caption{Normalized histograms of the power spectra that result from using inverse covariance weighting on PAPER-64 data after signal loss correction. Power spectrum points are computed from the peak of the distributions. Errors are computed using $95\%$ confidence intervals.}
	\label{fig:sigloss_datadist_inversecovariance}
\end{figure*}

@@

Investigating this further, the reason why the post-corrected power spectrum is so high is due to the small number of independent samples in PAPER-64. PAPER-64 has eight independent modes after fringe-rate filtering while our power spectrum bandwidth has $21$ channels. Consequently, the eigenvalues of $\hat{\textbf{C}}$ cannot possibly describe real fluctuations beyond the first $8$. Hence, there is a substantial down-weighting of many random modes, a large loss that the signal loss transfer function captures and corrects for. \cc{elaborate?} In the next section, we investigate alternate weighting schemes to inverse covariance weighting, with the goal of finding one that balances the aggressiveness of down-weighting contaminants with minimizing the loss of EoR.

@@

We saw in Section \ref{sec:otherweight} how limiting the number of down-weighted eigenmodes (i.e. flattening out part of the eigenspectrum) can help minimize signal loss. We experiment with this idea on PAPER-64 data, dialing the number of modes down-weighted from zero (which is equivalent to identity-weighting, or the unweighted case) to $21$ (which is full inverse covariance weighting of our $21$ channels). The power spectrum results for one $k$ value, both before and after signal loss correction, are shown in Figure \ref{fig:sigloss_modeloop}. We see that the amount of signal loss increases as weighting becomes more aggressive (grey curve). In other words, more random fluctuations are being overfit and subtracted as more modes are down-weighted. We also find that the power spectrum upper limit, post signal loss correction, increases with the number of down-weighted modes (black curve). As discussed previously, we are limited in our power spectrum accuracy by the number of independent samples in our dataset and our errors dramatically increase as we surpass the $\sim8$ independent samples that we have in fringe-rate filtered PAPER-64 data.

@@

\section{Acknowledgements}
The author would like acknowledge the UC Berkeley Chancellor's Fellowship, National Science Foundation Graduate Research Fellowship (Division of Graduate Education award 1106400), and thank Eric Switzer for helpful discussions. PAPER is supported by grants from the National Science Foundation (awards 0804508, 1129258, and 1125558). ARP, JCP, DCJ, and JEA would also like to acknowledge NSF support (awards 1352519, 1302774, 1401708, and 1455151, respectively). We also graciously thank SKA-SA for site infrastructure and observing support.
\label{sec:Ack}

>>>>

For our analysis, we use $8$ hours of LST (RA $0.5$-$8.6$ hours) and $51$ total baselines (\citet{ali_et_al2015} uses an RA range of $0$-$8.6$ hours). All power spectrum results are produced for a center frequency of $151$ MHz using a width of $10$ MHz ($20$ channels), identical to the analysis in \citet{ali_et_al2015}. We note that the PAPER-64 dataset that we use in this case study differs from that in \citet{ali_et_al2015} mainly by the fringe-rate filter. \cc{Danny says the next part is wrong, but that's what's in Zaki's paper... } In \citet{ali_et_al2015}, the applied filter was degraded by widening it in fringe-rate space. This was chosen in order to increase the number of independent modes and reduce signal loss. With the development of a robust method for assessing signal loss, we feel comfortable using a narrow filter --- the optimal fringe-rate filter --- in order to maximize sensitivity. This filter is computed for a fiducial $30$ m baseline at $150$ MHz, the center frequency in our band. The filter in both the fringe-rate domain and time domain is shown in Figure \ref{fig:frp}.

@@

We are interested in quantifying how much variance in $\textbf{e}$ is lost after weighting $\textbf{r}$ and estimating the power spectrum according to OQE formalism. We investigate this by comparing two quantities we call the input power spectrum and output power spectrum: $P_{in}$ and $P_{out}$, defined as (ignoring normalization factors):

@@

However, deeper investigations of these terms reveal that they contain non-negligible power. In fact, foreground modes and signal modes are anti-correlated on average, and this negative cross-term power arises from inverting the covariance matrix $\textbf{C}_{r}$. Mathematically,

\begin{eqnarray}
\textbf{C}_{r} &=& \langle \textbf{rr}^{\dagger} \rangle \\
&=& \langle \textbf{xx}^{\dagger} \rangle + \langle \textbf{xe}^{\dagger} \rangle + \langle \textbf{ex}^{\dagger} \rangle + \langle \textbf{ee}^{\dagger} \rangle,
\end{eqnarray}

\noindent where the data-only term (first term) and EoR-only term (last term) are approximately diagonal matrices. The middle two terms, however, have a larger amount of power in their off-diagonal terms relative to the diagonal elements, since $\textbf{e}$ and $\textbf{x}$ are statistically unrelated. Upon inversion, these off-diagonal terms become negative, driving the cross-terms in Equation \ref{eq:pout_expand} to have negative power. This effect is explained in more depth in \citet{switzer_et_al2015}.

@@

Our signal loss transfer function for one $k$ value is shown in Figure \ref{fig:sigloss_transfercurve} and compares the weighted case ($\textbf{R} = \textbf{C}^{-1}$, left plot) with the unweighted case ($\textbf{R} = \textbf{I}$, right plot). The figures depict a kernel density estimation of our input to output power spectrum transfer function. To sample the distribution, we run $40$ total injection levels, where the amplitude of \textbf{e} is increased in even increments on a lin-log grid and ranges from $P_{in} \sim -10^{13} \, mK^{2} (h^{-1} Mpc)^{3}$ (negative $P_{in}$ is injected when one copy of \textbf{e} in Equation \eqref{eq:Pin} is negative and the other is positive) to $P_{in} \sim10^{13} \, mK^{2} (h^{-1} Mpc)^{3}$. For each injection, we perform $20$ bootstraps over baselines, as explained in Section \ref{sec:Boot}, resulting in a total of $800$ data points on our $P_{in}$ vs. $P_{out}$ grid. We have verified that the transfer function is symmetric for negative and positive $P_{in}$'s, and so we fold all the values into positive ones to increase our signal-to-noise.

The solid black diagonal line in Figure \ref{fig:sigloss_transfercurve} shows unity-transfer, which we expect to occur for the unweighted case ($\textbf{R} = \textbf{I}$). Therefore, it is worth thinking about the spread, or scatter, around this black line that we see in the right plot (the colored `heat-map'). One might imagine that there should be one true $P_{out}$ value for every $P_{in}$, implying one well-determined signal loss factor for every data value. Focusing on the unweighted case alone, the scatter observed arises from the fact that the values of the cross-terms involving $\textbf{e}$ and $\textbf{x}$ in Equation \eqref{eq:pout_expand} change for different mock EoR signals, $\textbf{e}$. Since we draw different random $\textbf{e}$'s for every bootstrap, there is a range of $P_{out}$ values that results from the same $P_{in}$. This randomness causes the spread seen in the right plot of Figure \ref{fig:sigloss_transfercurve}, as well as most of the spread in the weighted case (left plot).

There is a second component of scatter that is less intuitive, only applies (at a low level) to the weighted case, and represents intrinsic scatter in signal loss. The origin of this scatter stems from how much the random EoR signal $\textbf{e}$ happens to look like the data $\textbf{x}$. For example, if we are extremely unlucky, one realization of $\textbf{e}$ would have the same shapes, or eigenvectors, as $\textbf{x}$. An empirically-derived covariance would then down-weight these shapes, destroying the entire EoR signal. On the other hand, the less that $\textbf{e}$ looks like $\textbf{x}$, the less signal loss that would result. The intrinsic scatter we can get is not a dominant factor in this case but it is important to correct for the fact that a particular $P_{out}$ value could arise from a range of $P_{in}$ values. We correct for this as explained in the next section.

Finally, the solid grey horizontal lines in Figure \ref{fig:sigloss_transfercurve} mark the peak of the data distribution $\hat{\textbf{p}}$ (the distribution is created from the $20$ bootstraps) for both the inverse covariance weighted and unweighted cases. One can eyeball that a data value of $10^{5} \, mK^{2} (h^{-1} Mpc)^{3}$, for example, would map approximately to a value of $\sim10^{8} \, mK^{2} (h^{-1} Mpc)^{3}$ using the transfer curve for the inverse covariance weighted case, implying a signal loss factor of $\sim1000$.

\begin{figure*}
	\centering
	\includegraphics[width=1\textwidth]{plots/sigloss_transfercurve.png}
	\caption{Signal loss transfer functions showing the relationship of $P_{in}$ and $P_{out}$, as defined by Equations \eqref{eq:Pin} and \eqref{eq:sigloss}. Kernel density estimations of the power spectrum transfer functions are shown as colored heat-maps for the cases of inverse covariance weighted PAPER-64 data (left) and unweighted data (right). The solid black diagonal line marks a perfect unity mapping, and the solid grey horizontal line denotes the peak of $\hat{\textbf{p}}$, the data distribution. From these plots, it is clear that inverse covariance weighting results in $\sim3$ orders of magnitude of signal loss for power spectrum values above $\sim$$2 \times 10^{4}$ $mK^{2} \, (h^{-1} \, Mpc)^{3}$, whereas the unweighted case does not exhibit loss.}
	\label{fig:sigloss_transfercurve}
\end{figure*}

One peculiar aspect of Figure \ref{fig:sigloss_transfercurve} is the fact that at low $P_{in}$ values it appears that we can have signal gain ($P_{out} > P_{in}$). This is unphysical in nature but caused due to the dominating cross-terms involving $\textbf{e}$ and $\textbf{x}$ in Equation \eqref{eq:pout_expand} once $\textbf{e}$ becomes small. As $\textbf{e}$ increases, we move into a regime where $P_{out} \sim P_{in}$, and then eventually into a regime where $P_{out} < P_{in}$ when $\textbf{e}$ is large enough to be destroyed if weighting the data using itself. Although we only show figures for one $k$ value, we note that the shape of the transfer curve is nearly identical for all $k$'s (though we treat each $k$ separately).

@@

\cc{can someone help me supplement the next 3 paragraphs with math?}

Beginning with smoothed probability distributions of the transfer function for the two cases (weighted and unweighted), defined as:

\begin{equation}
\cc{fill in},
\end{equation}

\noindent we fit a polynomial to each, representing mean signal loss functions. We then account for the second component of scatter as described previously by de-convolving out $P_{out, unweighted}$ from $P_{out, weighted}$ for each $P_{in}$ value:

\begin{equation}
\cc{fill in}.
\end{equation}

In other words, we only account for the extra scatter that the weighted case has over the unweighted case, which, as can be seen in Figure \ref{fig:sigloss_transfercurve}, is not large.

Equipped with our new signal loss curves, we multiply our data distribution $\hat{\textbf{p}}$ element-wise to all $P_{in}$ columns, treating positive and negative halves of the data distribution separately:

\begin{equation}
\cc{fill in}.
\end{equation}

We note that for PAPER-64, $\hat{\textbf{p}}$ is sometimes narrowly peaked at positive power spectrum values (signifying detections of power), and sometimes centered around zero, representing measurements that are consistent with zero.

Finally, we sum up all rows of $P_{out}$ where each row is a $P_{in}$ probability distribution allowed by our measurements:

\begin{equation}
\cc{fill in}.
\end{equation}

This results in a final $P_{in}$ distribution --- the distribution of our data as seen through the signal loss lens. We compute power spectrum points from the peak of the histograms, and power spectrum errors from $95\%$ confidence intervals.

%\begin{figure*}
%	\centering
%	\includegraphics[trim={1cm 0cm 1cm 1cm},width=1\textwidth]{plots/sigloss_datadist_inversecovariance.png}
%	\caption{Normalized histograms of the power spectra that result from using inverse covariance weighting on %PAPER-64 data after signal loss correction. Power spectrum points are computed from the peak of the distributions. Errors are computed using $95\%$ confidence intervals.}
%	\label{fig:sigloss_datadist_inversecovariance}
%\end{figure*}

@@

Investigating this further, the reason why the post-corrected power spectrum is so high is due to the small number of independent samples in PAPER-64. PAPER-64 only has $\sim$$8$ independent modes after fringe-rate filtering while our power spectrum bandwidth has $21$ channels. Consequently, the eigenvalues of $\hat{\textbf{C}}$ cannot possibly describe real fluctuations beyond the first eight. Hence, there is a substantial down-weighting of many random modes, a large loss that the signal loss transfer function captures and corrects for. We elaborate on this point in the next section, as well as investigate alternate weighting schemes to inverse covariance weighting, with the goal of finding one that balances the aggressiveness of down-weighting contaminants with minimizing the loss of EoR.

@@

We saw in Section \ref{sec:otherweight} how limiting the number of down-weighted eigenmodes (i.e. flattening out part of the eigenspectrum) can help minimize signal loss. We experiment with this idea on PAPER-64 data, dialing the number of modes down-weighted from zero (which is equivalent to identity-weighting, or the unweighted case) to $21$ (which is full inverse covariance weighting of our $21$ channels). The power spectrum results for one $k$ value, both before and after signal loss correction, are shown in Figure \ref{fig:sigloss_modeloop}. We see that the amount of signal loss increases as weighting becomes more aggressive (grey curve). In other words, more random fluctuations are being overfit and subtracted as more modes are down-weighted. We also find that the power spectrum upper limit, post signal loss correction, increases with the number of down-weighted modes (black curve). As discussed previously, we are limited in our power spectrum accuracy by the number of independent samples in our dataset and our errors dramatically increase as we surpass the $\sim8$ independent samples that we have in fringe-rate filtered PAPER-64 data \cc{elaborate?}.

@@

\section{Acknowledgements}
The author would like acknowledge the UC Berkeley Chancellor's Fellowship, National Science Foundation Graduate Research Fellowship (Division of Graduate Education award 1106400), and thank Eric Switzer for helpful discussions. PAPER and HERA are supported by grants from the National Science Foundation (awards 0804508, 1129258, 1125558, 1440343, and 1636646). ARP, JCP, DCJ, and JEA would also like to acknowledge NSF support (awards 1352519, 1302774, 1401708, and 1455151, respectively). We also graciously thank SKA-SA for site infrastructure and observing support.
\label{sec:Ack}
